Combining information from photometric and spectroscopic data:\\ Don't waste your time in spectrophotometric calibration!

### 2014-09-03 tex/ms.tex
put relevant figures back
figure for new model including gp_length
fixing for issue #6

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
the ``$\ast$'' operator is the pixel-wise multiplication operator,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to
the mean model $\mu$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In this formulation the the mean model $\mu$ will be effectively
multiplied by $e ^{f(\alpha)}$ ({\color{red} could we just add this to
the covariance matrix with a single long quadratic ?}).  This
logarithmic formulation requires that $d$ be everywhere positive, and
the uncertainties will be truncated to only allow positive values and
are therefore incorrect in detail.  However, at high signal to noise
the results should be unaffected. {\color{red} As I understand it,
additional `bad lines' in the mean model could be accounted for using
narrow quadratics in the covariance matrix, though their
interpretation would be hellish. The best way is probably to explictly
model them in the mean model}

@@

\item We should test on fake data, with known squiggly
  calibration applied, to demonstrate that we can infer the
  calibration vector and the physical parameters with no bias.

\item We infer the parameters of the spectrum of a real cluster that
  has been photometrically calibrated.

\item Next, we infer the parameters of the spectrum of the same
cluster \emph{without} spectrophotometric calibration, to see if we
infer physical parameters consistent with the calibrated case.  How
many/which figures to show?

\item  Then, we show select marginalized or joint PDFs for many
clusters calibrated/uncalibrated, to assess consistency.  Or we just
show comparisons of confidence intervals for the parameters of a
number of clusters.

\item We infer the parameters of the calibrated spectrum without
including the noise model, or presumably the photometry (since
reconciling photometry and spectroscopy is done by our noise model).
What changed, what did we get wrong without the noise model?
{\color{blue} Should we keep the $f(\alpha)$ and the photometry?  This
just has the potential to show shortcomings in the spectrophotometric
calibration calibration}

@@

The final mean model is then given by 
\begin{eqnarray}\label{eq:MeanModel} 
\mu(\theta) &  = & (G_\sigma  \otimes F_{\theta^\prime})(\lambda) +
\sum_{\ell=0}^{N_\ell} S_\ell \\
S_\ell & = & \frac{A_\ell}{\sigma_e\sqrt{2\pi}} \exp(-\frac{(\lambda -
 \lambda_\ell)^2}{2\sigma_e^2})\\
\theta & \equiv & \left[ \{\theta^\prime\}, \sigma, \{A_\ell\}_{\ell=0}^{N_l},
  \sigma_e \right]
\quad ,
\end{eqnarray}
where $G_\sigma$ represents a Gaussian of unit area 
and standard deviation $\sigma$, 
$\otimes$ indicates convolution, 
and the full parameter set $\theta$ now includes
the instrumental wavelength broadening $\sigma$ 
and the dispersion $\sigma_e$ and
amplitudes $A_\ell$ of the $N_l$ nebular emission and interstellar
absorption lines.
This is the mean model which appears in equation
\ref{eq:spectroscopicLF}.

@@

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/pfig_b192-g242_225_cal.png}
\caption{ Model SEDs.  Top: The model photometric SED resulting from
draws from the posterior parameter space conditioned ({\it cyan}) on
the observed photometry ({\it magenta}) and spectra.  Bottom:
uncertainty normalized residuals of the models from the observations.
\label{fig:inferred_sed}}
\end{figure}

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/ptri_b192-g242_225_cal.png}
\caption{ Marginalized and joint PDFs of subset of the physical
parameters $\theta$ inferred from the spectophotometrically calibrated
spectrum of \excluster. The blue lines indicate the parameter values
that resulted from the powell minimization.
\label{fig:inferred_params}}
\end{figure}

@@

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/cfig_b192-g242_225.png}
\caption{Left: Samples of the effective multiplicative calibration
vector for the spectrophotometrically calibrated spectrum of
\excluster~ ({\it cyan}) compared to that obtained from ancillary
photometric data ({\it magenta}).  
%Right: The same, but for the uncalibrated data.
 \label{fig:inferred_calib}}
\end{figure*}

@@

Our strategy is easily extensible to more complicated models, which
include for example night sky-line emission in the spectroscopy (or
residuals from an intial sky-line subtraction), and uncertainties in
wavelength calibration. Furthermore, more complex physical models may
be considered for the mean object model, for example including SFH
parameters so as to model galaxies.  {\color{blue} DISCUSS MORE
COMPLICATED COVARIANCE KERNELS, MULTIPLE KERNELS, etc.  MODEL
CHECKING.}

@@

Limitations.  The strategy we have advocated has several limitations.
First, it requires a generative model for the spectrum of the object
of interest, and is thus ill-suited to discovery projects.  Second,
care must be taken to avoid absorbing interesting model deficiencies
into the calibration model. For example, in the model we have
described, it is possible that broad shallow features (for example due
to molecular bands or absorption line complexes) will be absorbed into
the Gaussian process. Third, we have made the assumption that the
spectroscopy and the photometry are sampling the same populations.  In
practice, differences between spectroscopic and photometric apertures
often mean that the photometry and the spectroscopy are sampling
slightly different physical regions.  The correct response to this
situation is to explicitly model the differences, incorporating prior
information about the spatial variation of physical properties.
Finally, this method is computationally intensive and expensive, since
we are constantly inverting largish matrices.

{\color{blue} BDJ - This paragraph is probably too calibration
focused. We want to focus more on spec + phot combinations.  At least,
this paragraph should not be in the discussion section.}
Fundamentally, calibration involves observing a `known' or `truth'
with an instrument, and using the observation to infer the effects of
the instrument.  However, we do not have knowledge of a single truth
for any object, only relative probabilities for what might be true.
In this sense, one should never use a single point estimate of the
calibration, the full distribution of the possible calibrations should
be inferred from our probabilistic knowledge of the calibration object
and forward propogated \citep[e.g.][]{lee11}. Once this is done, the only the
only distinction between 'calibration' objects and targets of interest
is the complexity of the model for the object and the level of
uncertainty in the calibration object spectrum.  If one is interested
in inferring parameters of interest of spectral models of
astrophysical objects, then there is no reason not to use those
objects themselves as calibration objects.

>>>>

We take the calibration-vector that modifies the mean model to produce
the observed spectrum as multiplicative with respect to the mean
model.  In this case we define the residual of the data from the model
in terms of the logarithmic differences.  This logarithmic formulation
requires that $d$ be everywhere positive, and distorts the
uncertainties, especially at low S/N ratio.  The advantages of
defining the residual in this way are that the model remains linear
(and therefore fast) and the calibration vector is forced to be
positive everywhere.  However, at high signal to noise the distortion
of the uncertainties will be minimal.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelengths scale and
amplitude of the residual calibration issues,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to
the mean model $\mu$. In this formulation the the mean model $\mu$ will be effectively
multiplied by $e ^{f(\alpha)}$ 
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

@@

\item We should test on fake data, with known squiggly
  calibration applied, to demonstrate that we can infer the
  calibration vector and the physical parameters with no bias.  We
  should use the squiggly calibration derived from one of Nelson's
  calibrated/uncalibrated spectra.  We should also do this for
  differnt S/N to find where the logarithmic residual formulation
  breaks down.

\item We infer the parameters of the spectrum of a real cluster that
has been photometrically calibrated. Then, we infer the parameters of
the spectrum of the same cluster \emph{without} spectrophotometric
calibration, to see if we infer physical parameters consistent with
the calibrated case.  How many/which figures to show?

\item  Then, we show select marginalized or joint PDFs for many
clusters calibrated/uncalibrated, to assess consistency.  Or we just
show comparisons of confidence intervals for the parameters of a
number of clusters, calibrated/uncalibrated.

\item We infer the parameters of the calibrated spectrum without
including the noise model, or presumably the photometry (since
reconciling photometry and spectroscopy is done by our noise model).
What changed, what did we get wrong without the noise model?
{\color{blue} Should we keep the $f(\alpha)$ and the photometry?  This
just has the potential to show shortcomings in the spectrophotometric
calibration}

@@

The final mean model is then given by 
\begin{eqnarray}\label{eq:MeanModel} 
\mu(\theta) &  = & (G_\sigma  \otimes F_{\theta^\prime})(\lambda) +
\sum_{\ell=0}^{N_\ell} S_\ell(\lambda) \\
S_\ell (\lambda) & = & \frac{A_\ell}{\sigma_e\sqrt{2\pi}} \exp(-\frac{(\lambda -
 \lambda_\ell)^2}{2\sigma_e^2})\\
\theta & \equiv & \left[ \{\theta^\prime\}, \sigma, \{A_\ell\}_{\ell=0}^{N_l},
  \sigma_e \right]
\quad ,
\end{eqnarray}
where $G_\sigma$ represents a Gaussian of unit area 
and standard deviation $\sigma$, 
$\otimes$ indicates convolution, 
and the full parameter set $\theta$ now includes
the instrumental wavelength broadening $\sigma$ 
and the dispersion $\sigma_e$ and
amplitudes $A_\ell$ of the $N_l$ nebular emission and interstellar
absorption lines.
This is the mean model which appears in equation
\ref{eq:spectroscopicLF}.

@@

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{figures/pfig_b192-g242_225_cal.png}
\caption{ Model SEDs.  Top: The model photometric SED resulting from
draws from the posterior parameter space conditioned ({\it cyan}) on
the observed photometry ({\it magenta}) and spectra.  Bottom:
uncertainty normalized residuals of the models from the observations.
\label{fig:inferred_sed}}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{figures/ptri_b192-g242_225_cal.png}
\caption{ Marginalized and joint PDFs of subset of the physical
parameters $\theta$ inferred from the spectophotometrically calibrated
spectrum of \excluster. The blue lines indicate the parameter values
that resulted from the powell minimization.
\label{fig:inferred_params}}
\end{figure}

@@

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/cfig_b192-g242_225.png}
\caption{Left: Samples of the effective multiplicative calibration
vector for the spectrophotometrically calibrated spectrum of
\excluster~ ({\it cyan}) compared to that obtained from ancillary
photometric data ({\it magenta}).  
%Right: The same, but for the uncalibrated data.
 \label{fig:inferred_calib}}
\end{figure*}

@@

Our strategy is easily extensible to more complicated models, which
include for example night sky-line emission in the spectroscopy (or
residuals from an intial sky-line subtraction), and uncertainties in
wavelength calibration. Furthermore, more complex physical models may
be considered for the mean object model, for example including SFH
parameters so as to model galaxies.  {\color{blue} DISCUSS MORE
COMPLICATED COVARIANCE KERNELS, MULTIPLE KERNELS, etc.  MODEL
CHECKING. As I understand it, additional `bad lines' in the mean model
could be accounted for using narrow quadratics in the covariance
matrix, though their interpretation would be hellish. The best way is
probably to explictly model them in the mean model}

@@

Limitations.  The strategy we have advocated has several limitations.
First, it requires a generative model for the spectrum of the object
of interest, and is thus ill-suited to discovery projects.  Second,
care must be taken to avoid absorbing interesting model deficiencies
into the calibration model. For example, in the model we have
described, it is possible that broad shallow features (for example due
to molecular bands or absorption line complexes) will be absorbed into
the Gaussian process. Third, we have made the assumption that the
spectroscopy and the photometry are sampling the same populations.  In
practice, differences between spectroscopic and photometric apertures
often mean that the photometry and the spectroscopy are sampling
slightly different physical regions.  The correct response to this
situation is to explicitly model the differences, incorporating prior
information about the spatial variation of physical properties.
Fourth, the method will break down at very low S/N due to the
logarithmic formulation of the residual in
equation~(\ref{eq:spectroscopicLF}).  Finally, this method is
computationally intensive and expensive, since we are constantly
inverting largish matrices.

{\color{blue} BDJ - This paragraph is probably too calibration
focused. We want to focus more on spec + phot combinations.  At least,
this paragraph should not be in the discussion section.}
Fundamentally, calibration involves observing a `known' or `truth'
with an instrument, and using that observation to infer the effects of
the instrument.  However, we do not have knowledge of a single truth
for any object, only relative probabilities for what might be true.
In this sense, one should never use a single point estimate of the
calibration, the full distribution of the possible calibrations should
be inferred from our probabilistic knowledge of the calibration object
and forward propogated \citep[e.g.][]{lee11}. Once this is done, the
only distinction between 'calibration' objects and targets of interest
is the complexity of the model for the object and the level of
uncertainty in the calibration object spectrum.  If one is interested
in inferring parameters of interest of spectral models of
astrophysical objects, then there is often no reason not to use those
objects themselves as calibration objects.
